---
title: "Rationale for MASSExtra"
author: "Bill Venables"
date: "`r Sys.Date()`"
output:
  html_vignette: null
  pdf_document:
    includes:
      in_header: header.tex
  html_document:
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Rationale for MASSExtra} 
  %\VignetteEncoding{UTF-8} 
  %\VignetteEngine{knitr::rmarkdown}
---

```{r presetup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "",
                      message = FALSE,
                      warning = FALSE)
old_plot_new <- getHook("plot.new")
old_before_plot_new <- getHook("before.plot.new")
setHook("plot.new", list(), "replace")
setHook("before.plot.new", rev(old_before_plot_new)[1], "replace")
old_par <- par(no.readonly = TRUE)
old_options <- options()

library(tidyverse)
old_theme <- theme_get()
```

# Preamble

This extension package to the classical `MASS` package (Venables &
Ripley, of ancient lineage), whose origins go back to nearly 30 years,
comes about for a number of reasons.

Firstly, in my teaching I found I was using some of the old functions
in the package with consistently different argument settings to the
defaults.  I was also interested in supplying various convenience
extensions that simplified teaching and including various tweaks to
improve the interface.  Examples follow below.

Secondly, I wanted to provide a few functions that were mainly useful
as programming examples.  For example, the function `zs` and its
allies `zu`, `zq` and `zr` are mainly alternatives to `base::scale`,
but they can be used to show how to write functions that can be used
in fitting models in such a way that they work as they should when the
fitted model object is used for prediction with new data.

### Masking `select` from other packages

Finally, there is the perennial `select` problem.  When `MASS` is used
with other packages, such as `dplyr` the `select` function can easily
be masked, causing confusion for users.  `MASS::select` is rarely
used, but `dplyr::select` is fundamental.  There are standard ways of
managing this kind of masking, but what we have done in `MASSExtra` is
to export the more common functions used from `MASS` along with the
extensions, in such a way that users will not need to have `MASS`
attached to the search path at all, and hence masking is unlikely.

The remainder of this document will do a walk-through of some of the
new functions provided by the package.  We begin by setting the
computational context:

```{r packages_etc}
suppressPackageStartupMessages({
  library(visreg)
  library(knitr)
  library(tidyverse)
  library(patchwork)
  library(MASSExtra)
})
options(knitr.kable.NA = "")
theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5)))
```


# Amble

We now consider some of the extensions that the package offers to the
originals.  Most of the extensions will have a name that includes an
underscore of two somewhere to distinguish it from the V&R original.
Note that the original version is _also_ exported so that scripts that
use it may do so without change, via the new package.

## The `box_cox` extensions

This original version, `boxcox` has a fairly rigid display for the
plotted output which has been changed to give a more easily
appreciated result.  The $y-$axis has been changed to give the
likelihood-ratio statistic rather than the log-likelihood, and for the
$x-$axis some attempt has been made to focus on the crucial region for
the transformation parameter, $\lambda$,

The following example shows the old and new plot versions for a simple
example.

```{r setup, fig.height = 5, fig.width = 10, out.width="100%", fig.cap="Box-cox, old and new displays"}
par(mfrow = c(1, 2))
mod0 <- lm(MPG.city ~ Weight, Cars93)
boxcox(mod0)  ## MASS
box_cox(mod0) ## MASSExtra tweak
```

In addition, there are functions `bc` to evaluate the transformation
for a given exponent, and a function `lambda` which finds the optimum
exponent (not that a precise exponent will usually be needed).

It is interesting to see how in this instance the transformation can
both straighten the relationship and provide a scale in which the
variance is more homogeneous.  See Figure 2.

```{r, fig.height=5, fig.width=10, out.width="100%", fig.cap="The Box-Cox transformation effect"}
p0 <- ggplot(Cars93) + aes(x = Weight) + geom_point(colour = "#2297E6")  + xlab("Weight (lbs)") +
  geom_smooth(se = FALSE, method = "loess", formula = y ~ x, size=0.7, colour = "black")
p1 <- p0 + aes(y = MPG.city) + ylab("Miles per gallon (MPG)") + ggtitle("Untransformed response")
p2 <- p0 + aes(y = bc(MPG.city, lambda(mod0))) + ggtitle("Transformed response") + 
  ylab(bquote(bc(MPG, .(round(lambda(mod0), 2)))))
p1 + p2
```

A more natural scale to use, consistent with the Box-Cox suggestion,
would be the reciprocal.  For example we could use $\mbox{GPM} =
100/\mbox{MPG}$ the "gallons per 100 miles" scale, which would have
the added benefit of being more-or-less what the rest of the world
uses to gauge fuel efficiency outside the USA.  Readers should try
this for themselves.

## Stepwise model building extensions

The primary `MASS` functions for refining linear models and their
allies are `dropterm` and `stepAIC`.  The package provides a few
extensions to these, but mainly a change of defaults in the argument
settings.

1. `drop_term` is a front-end to `MASS::dropterm` with a few tweaks.
   By default the result is arranged in sorted order, i.e. with
   `sorted = TRUE`, and also by default with `test = TRUE` (somewhat
   in defiance of much advice to the contrary given by experienced
   practitioners: _caveat emptor!_).
   
   The user may specify the test to use in the normal way, but the
   default test is decided by an ancillary generic function,
   `default_test`, which guesses the appropriate test from the object
   itself.  This is an S3 generic and further methods can be supplied
   for new fitted model objects.
   
   There is also a function `add_term` which provides similar
   enhancements to those provided by `drop_term`.  In this case, of
   course, the consequences of _adding_ individual terms to the model
   are displayed, rather than of _dropping_ them.  It follows that
   using `add_term` you will always need to provide a scope
   specification, that is, some specification
   of what extra terms are possible additions.  

   In addition `drop_term` and `add_term` return an object which
   retains information on the criterion used, `AIC`, `BIC`, `GIC` (see
   below) or some specific penalty value `k`.  The object also has a
   class `"drop_term"` for which a `plot` method is provided.  Both
   the `plot` and `print` methods display the criterion.  See the
   example below for how this is done.
   
1. `step_AIC` is a front-end to `MASS::stepAIC` with the default
   argument `trace = FALSE` set.  This may of course be over-ruled,
   but it seems the most frequent choice by users, anyway.  In
   addition the actual criterion used, by dafault `k = 2`, i.e. AIC,
   is retained with the result and passed on to methods in much the
   same say as for `drop_term` above.
   
   Since the (default) criterion name is encoded in the function name,
   two further versions are supplied, namely `step_BIC` and `step_GIC`
   (again, see below), which use a different, and obvious, default
   criterion.
   
   In any of `step_AIC`, `step_BIC` or `step_GIC` a different value of
   `k` may be specified in which case that value of `k` is retained
   with the object and displayed as appropriate in further methods.
   
   Finally in any of these functions `k` may be specified either as a
   numeric penalty, such as `k = 4` for example, or by character
   string `k = "AIC"` or `k = "BIC"` with an obvious meaning in either
   case.
   
1. Criteria.  The __Akaike Information Criterion__, AIC, corresponds
   to a penalty `k = 2` and the __Bayesian Information Criterion__,
   BIC, corresponds to `k = log(n)` where `n` is the sample size.  In
   addition to these two the present functions offer an intermediate
   default penalty `k = (2 + log(n))/2` which is "not too strong and
   not too weak", making it the __Goldilocks Information Criterion__,
   GIC.  There is also a standalone function `GIC` to evaluate this
   `k` if need be.
   
   This suggestion appears to be original, but _no particular claim is
   made for it_ other than with intermediate to largish data sets it
   has proved useful for exploratory purposes in our
   experience.  
   
   Our strong advice is that these tools should __*only*__ be used for
   exploratory purposes in any case, and should __*never*__ be used in
   isolation.  They have a well-deserved very negative reputation when
   misused, as they commonly are.
   
## Examples

We consider the well-known (and much maligned) Boston house price
data.  See `?Boston`.  We begin by fitting a model that has more terms
in it than the usual model, as it contains a few extra quadratic
terms, including some key linear by linear interactions.

```{r, fig.width=8, fig.height=6, fig.align="center", out.width="75%"}
big_model <- lm(medv ~ . + (rm + tax + lstat + dis)^2 + poly(dis, 2) + poly(rm, 2) +
                   poly(tax, 2) + poly(lstat, 2), Boston)
big_model %>% drop_term(k = "GIC") %>% plot() %>% kable(booktabs=TRUE, digits=3)
```

Unlike `MASS::dropterm`, the table shows the terms beginning with the
most important ones, that is those which, if dropped, would _increase_
the criterion and ending with those of least looking importance, that
is those whose removal would most _decrease_ the criterion.  And also
note that here we are using the `GIC`, which is displayed in the
output.

Note particularly that rather than give the _value_ of the criterion
by default the table and plot show _change_ in the criterion which
would result if the term is removed from the model at that point.
This is a more meaningful quantity, and invariant with respect to the
way in which the log-likelihood is defined.

The `plot` method gives a graphical view of the same key bits of
information, in the same vertical order as given in the table.  Terms
whose removal would (at this point) improve the model are shown in
_red_ and those which would not, and hence should (again, at this
point) be retained are shown in _blue_.

With all stepwise methods it is critically important to notice that
the whole picture can change once any change is made to the current
model.  This terms which appear "promising" at this stage may not seem
so once any variable is removed from the model or some other variable
brought into it.  This is a notoriously tricky area for the
inexperienced.

Notice that the `plot` method returns the original object, which can
then be passed on via a pipe to more operations.  (`kable` does not,
so this pipe sequence cannot be changed.)

We now consider a refinement of this model by stepwise means, but
rather than use the large model as the starting point, we begin with a
more modest one which has no quadratic terms.

```{r, fig.width=8, fig.height=6, fig.align="center", out.width="75%"}
base_model <- lm(medv ~ ., Boston)
gic_model <- step_GIC(base_model, scope = list(lower = ~1, upper = formula(big_model)))
drop_term(gic_model) %>% plot() %>% kable(booktabs = TRUE, digits = 3)
```

The model is likely to be over-fitted.  To follow up on this we could
look at profiles of the fitted terms as an informal way of model
'criticism'.

```{r, fig.width=14, fig.height=6, fig.align="center", out.width="100%", fig.cap="Two component visualisations for the Boston house price model"}
capture.output(suppressWarnings({
   g1 <- visreg(gic_model, "dis", plot = FALSE)
   g2 <- visreg(gic_model, "lstat", plot = FALSE)
   plot(g1, gg = TRUE) + plot(g2, gg = TRUE) 
})) -> void
```

The case for curvature appears to be fairly weak, in each case with
departure from a straight line dependence depending on a relatively
few observations with high values for the predictor.  (Notice how hard
you have to work to prevent `visreg` from generating unwanted output.)

As an example of `add_term`, consider going from what we have called
the `base_model` to the `big_model`, or at least what might be the
initial step:

```{r, fig.width=8, fig.height=6, fig.align="center", out.width="75%"}
add_term(base_model, scope = formula(big_model), k = "gic") %>% 
   plot() %>% kable(booktabs = TRUE, digits = 3)
```

So in this case, your best first step would be to _add_ the term which
most decreases the criterion, that is, the one nearest the bottom of
the table (or display).

### Non-gaussian models

For a non-gaussian model consider the Quine data (`?quine`) example
discussed in the _MASS_ book.  We begin by fitting a full negative
binomial model and refine it using a stepwise algorithm.

```{r}
quine_full <- glm.nb(Days ~ Age*Eth*Sex*Lrn, data = quine)
drop_term(quine_full) %>% kable(booktabs = TRUE, digits = 4)
quine_gic <- step_GIC(quine_full)
drop_term(quine_gic) %>% kable(booktabs = TRUE, digits = 4)
```

So GIC refinement has led to the same model as in the _MASS_ book,
which in more understandable form would be written `Days ~ Sex/(Age +
Eth*Lrn)`.  Note also that the default test is in this case
the likelihood ratio test.  

For a different example, consider an alternative way to model the MPG
data, rather than transforming to an inverse scale, using a
generalized linear model with an inverse link and a gamma response.

```{r}
mpg0 <- glm(MPG.city ~ Weight + Cylinders + EngineSize + Origin, 
            family = Gamma(link = "inverse"), data = Cars93)
drop_term(mpg0) %>% kable(booktabs = TRUE, digits = 3)
mpg_gic <- step_down(mpg0, k = "gic")      ## simple backward elimination, mainly used for GLMMs
drop_term(mpg_gic) %>% kable(booktabs = TRUE, digits = 3)
```

We can see something of how well the final model is performing by
looking at a slightly larger model fitted on the fly in `ggplot`:

```{r, fig.width=10, fig.height=7, fig.align="center", out.width="80%", fig.cap="MPG model using a Gamma family with inverse link"}
ggplot(Cars93) + aes(x = Weight, y = MPG.city, colour = Cylinders) + geom_point() +
   geom_smooth(method = "glm", method.args = list(family = Gamma), formula = y ~ x) +
   ylab("Miles per gallon (city driving)") + scale_colour_brewer(palette = "Dark2")
```

The function `step_down` is a simple implementation of backward
elimination with the sole virtue that it works for (Generalised)
Linear Mixed Models, or at least for the fixed effect component of
them, whereas other stepwise methods do not (yet).  As fitting GLMMs
can be very slow, going through a full stepwise process could be very
time consuming in any case.

## Scaling functions

When a function such as `base::scale`, `stats:poly` or `splines:ns`
(also exported from `MASSExtra`) is used in modelling it is important
that the fitted model object has enough information so that when it is
used in prediction for new data, the same transformation can be put in
place with the new predictor variable values.  Setting up functions in
such a way to enable this is a slightly tricky exercise.  It involves
writing a method function for the S3 generic function
`stats::makepredictcall`.  To illustrate this we have supplied four
simple functions that
employ the technique.  

* `zs` ("z-score") is essentially the same as `base::scale` with the default argument settings,
* `zu` allows re-scaling to a fixed range of [0, 1], often used in neural network models,
* `zq` allows a quantile scaling where the location is the lower quartile and the scale is the inter-quartile range. 
   In other words, the scaling is to [0,1] _within the box_ of a boxplot.  _(Go figure.)_
* `zr` allows a "robust" scaling where the location is the median and the scale uses `stats::mad`

The only real interest in these very minor convenience functions lies
in how they are programmed.  See the code itself for more details.

<!-- The following chain from the package code -->
<!-- sets out the steps. -->

<!-- ```{r} -->
<!-- dput(MASSExtra:::makepredictcall.normalise) ## exported only as an S3 method -->
<!-- dput(.normalise)                            ## an exported helper, but with a hidden name -->
<!-- dput(zs)                                    ## the scale() equivalent -->
<!-- dput(zq)                                    ## the 'boxplot scaling' (based on quantiles) -->
<!-- ``` -->

## Kernel density estimation

Release 1.1.0 contains two functions for kernel density estimation: one- and two-dimensional.

* `kde_1d` offers a similar functionality to `stats::density`, though with two additional
  features that may be useful in some situations, namely
  - The kernel function may be specified as an R function, as well as as a character
    string to select from a preset list.  Users wishing to write a special kernel function
    should do so in line with, for example `demoKde::kernelBiweight` or `demoKde::kernalGaussian`,
    using the same argument list with the same intrinsic meanings for the arguments themselves.
  - The kernel density estimate may be "folded" to emulate the effect of fitting a density with
    a known finite range for the underlying distribution.  This amounts to fitting the kde initially
    with unrestricted range and "folding back" the parts beyond the known range, adding them
    on to the mirror image components inside the range.  This strategy appears to give a credible
    result, though no particular claim is made for it on theoretical grounds.  See the examples.

* `kde_2d` uses much the same computational ideas as in `MASS::kde2d` (due to Prof. Brian Ripley),
  but uses an approximation that allows the algorithm to scale much better for both large data sets
  and large resolution in the result.  Indeed the approximation improves as the resolution increases,
  so the default size is now $512\times512$ rather than $25\times25$ as it is for `MASS::kde2d`.  This
  function also allows the kernel function(s) to be either specified or user-defined, as for
  `kde_1d` above.  Folding is not implemented, however.
  
Both functions produce objects with a class agreeing with the name of the calling function, and
suitable `plot` and `print` methods are provided.

Two examples follow.  The first shows (mainly) the surprising capacity for a log-transformation
to amplify what is essentially a trivial effect into something that appears impressive!

```{r, fig.width=10, fig.height=7, fig.align="center", out.width="80%", fig.cap="A one-dimensional KDE of a predictor variable from the Boston data set"}

Criminality <- with(Boston, log(crim))
kcrim <- kde_1d(Criminality, n = 1024, kernel = demoKde::kernelBiweight)
kcrim
plot(kcrim)
```

We now take this further into a two-dimensional example

```{r, fig.width=10, fig.height=7, fig.align="center", out.width="80%", fig.cap="A two-dimensional KDE using two predictors from the Boston data set"}
Spaciousness <- with(Boston, sqrt(rm))
kcrimrm <- kde_2d(Criminality, Spaciousness, n = 512, kernel = "opt")
kcrimrm
plot(kcrimrm, ## col = hcl.colors(25, rev = TRUE),
     xlab = expression(italic(Criminality)),
     ylab = expression(italic(Spaciousness)))
contour(kcrimrm, col = "dark green", add = TRUE)
```

An even more deceptive plot uses `persp`:
```{r, fig.width=10, fig.height=7, fig.align="center", out.width="80%", fig.cap="A two-dimensional KDE presented as a perspective plot"}
with(kcrimrm, persp(x, 10*y, 3*z, border="transparent", col = "powder blue",
                    theta = 30, phi = 15, r = 100, scale = FALSE, shade = TRUE, 
                    xlab = "Criminality", ylab = "Spaciousness", zlab = "kde"))
```



# Postamble

In this final section we mainly give a list of functions provided by
the package, and their origins.

We begin by giving a list of functions in the `MASS` package which are
_not_ re-exported from the `MASSExtra` package.  If you need any of
these you will need either to attach the `MASS` package itself, or use
the qualified form `MASS::<name>`.

```{r, echo=FALSE}
setdiff(getNamespaceExports("MASS"), getNamespaceExports("MASSExtra")) %>% sort() %>%  noquote()
```

The following objects _are_ re-exported from the `MASSExtra` package,
and hence may be used directly, if needed.

```{r, echo=FALSE}
intersect(getNamespaceExports("MASS"), getNamespaceExports("MASSExtra")) %>% sort() %>%  noquote()
```

The following functions are _new_ to the `MASSExtra` package, some of
which are obviously refinements of their `MASS` workhorse counterparts.

```{r, echo=FALSE}
setdiff(getNamespaceExports("MASSExtra"), getNamespaceExports("MASS")) %>% 
   setdiff(getNamespaceExports("splines")) %>% 
   grep("^[.]__T", ., invert = TRUE, value = TRUE) %>% ## exclude S4 class objects
   sort() %>% noquote()
```

Finally the following objects are re-exported from `splines`:


```{r, echo=FALSE}
intersect(getNamespaceExports("MASSExtra"), getNamespaceExports("splines")) %>% sort() %>%  noquote()
```


Only four of the `MASS` data sets are included in `MASSExtra`, namely
`Cars93`, `Boston`, `quine` and `whiteside`.  Other data sets from
`MASS` itself will need to be accessed directly, e.g. `MASS::immer`.

```{r cleanup, include=FALSE}
setHook("plot.new", old_plot_new, "replace")
setHook("before.plot.new", old_before_plot_new, "replace")
par(old_par)
options(old_options)
theme_set(old_theme)
```

